---
title: "Project 2 - Mental Health Dataset"
author: "Theresa Benny"
format: html
editor: visual
---

## Approach

The first dataset contains simulated mental health survey data collected from participants across several cities in the United States. Each row represents a participant and includes demographic information, therapy participation, medication usage, screening scores, and monthly assessment data. However, the dataset is currently stored in a **wide and inconsistent format**, making it difficult to analyze directly.

Several columns contain combined variables (such as screening scores), inconsistent missing value labels (e.g., "NA", "missing", "-", and "n/a"), and repeated measures across months (January through May). The goal of this dataset transformation is to convert the data into a **tidy structure** where each variable forms its own column and each observation forms its own row. This tidy format will allow easier analysis of trends in mental health screening scores, therapy participation, and stress levels across participants.

To preserve the original dataset structure, I will first recreate the dataset in its **original wide format** and store it as a `.csv` file. This raw file will be committed to my GitHub repository before performing any transformations.

The raw dataset contains columns such as:

-   Participant demographics (AgeGroup, Gender, City_State)

-   Mental health screening scores (PHQ-9 and GAD-7)

-   Monthly measurement columns (Jan–May)

-   Therapy and medication information

-   Diagnostic status

-   Lifestyle factors (WorkHours, Sleep, StressScale)

-   Insurance type and survey metadata

This file will serve as the **unmodified input dataset** for the tidying pipeline.

The dataset will be imported into R using `read_csv()` from the **tidyverse** package. I will inspect the structure of the dataset using functions such as `glimpse()` and `summary()` to understand the data types and identify inconsistencies.

Column names will be standardized using `rename_with()` to ensure consistent formatting (e.g., lowercase variable names and underscores instead of spaces). This improves readability and ensures compatibility with downstream analysis workflows.

The `ScreeningScores` column contains two separate measurements in the format:

PHQ-9:score, GAD-7:score

These represent two different mental health screening scales for depression and anxiety. I plan to separate this column into two variables using `separate()` from **tidyr**.

The dataset currently contains multiple columns representing monthly measurements:

-   Jan

-   Feb

-   Mar

-   Apr

-   May

These columns represent repeated observations over time and should be converted into a tidy structure using `pivot_longer()`.

The dataset contains several different representations of missing data, including:

-   "NA"

-   "missing"

-   "-"

-   "n/a"

These values will be standardized into proper `NA` values using `na_if()` or `mutate()` operations. This ensures that missing data is handled consistently during analysis.

Some numeric fields currently include text (for example, "33 hours" or "7.8 hours"). These variables will be cleaned using string manipulation functions such as `str_remove()` before converting them to numeric types.

Once the dataset has been converted into tidy format, I will perform exploratory analysis to understand patterns in mental health outcomes.

Potential analyses include:

Comparing **PHQ-9 and GAD-7 scores across age groups**

Examining whether **therapy type is associated with lower stress scores**

Analyzing **monthly changes in mental health screening results**

Exploring relationships between **sleep hours, work hours, and stress levels**

Summary tables will be generated using `dplyr` functions such as `group_by()` and `summarise()`. Visualizations will be created using **ggplot2**, including bar charts and line plots showing mental health trends over time.

Several challenges are expected when tidying this dataset:

1.  **Inconsistent missing value labels** that must be standardized.

2.  **Combined variables** in the screening scores column that require separation.

3.  **Monthly measurements stored across multiple columns**, which require reshaping.

4.  **Text embedded within numeric variables** (e.g., "hours") that must be cleaned before analysis.

Addressing these issues will ensure the final dataset follows tidy data principles and is suitable for reproducible analysis.

## 

## Codebase

```{r}

library(tidyverse)

#import mental health data
mental_raw <- read_csv("untidy_mental_health_data.csv")

glimpse (mental_raw)
```

```{r}
#Preview the raw data

head(mental_raw)


```

From this preview, the dataset appears untidy for several reasons. Monthly observations are spread across multiple columns rather than stored in a single month variable. Some columns also contain inconsistent text formatting, and missing values may be represented in different ways instead of using standard NA values. In addition, certain variables may need to be separated or cleaned before they can be analyzed properly.

```{r}
#Let's check the column names now
names(mental_raw)

```

Reviewing the column names helps identify which variables are participant characteristics and which columns represent repeated monthly measurements. It also helps determine whether variable names should be standardized to improve readability and consistency in the tidied dataset.

```{r}
# Check for Missing or Inconsistent Values

summary(mental_raw)
```

```{r}
#Let's start cleaning the dataset by renaming all the column names consistently

mental_clean <- mental_raw %>%
  rename_with(tolower)

glimpse(mental_clean)



```

```{r}
#Reshape monthly data to long format

mental_tidy <- mental_clean %>%
  pivot_longer(
    cols = jan:may,
    names_to = "month",
    values_to = "monthly_value"
  )
glimpse(mental_tidy)
```

```{r}
#Next let's handle missing and inconsistent values
mental_tidy <- mental_tidy %>%
  mutate(across(where(is.character), ~ na_if(.x, "missing"))) %>%
  mutate(across(where(is.character), ~ na_if(.x, "-"))) %>%
  mutate(across(where(is.character), ~ na_if(.x, "n/a")))

glimpse(mental_tidy)

#Cleaning worked!
```

```{r}
#Screening scores column has multiple value packed into it, so let's split those out. 
mental_tidy %>%
  select(screeningscores) %>%
  distinct() %>%
  head(10)

#PHQ and GAD scores are formatted differently

mental_tidy <- mental_tidy %>%
  mutate(
    phq9_score = str_extract(screeningscores, "(?<=PHQ-9=)\\d+|(?<=PHQ9:)\\d+"),
    gad7_score = str_extract(screeningscores, "(?<=GAD-7=)\\d+|(?<=GAD7:)\\d+")
  )

mental_tidy %>%
  select(screeningscores, phq9_score, gad7_score) %>%
  head(10)

mental_tidy <- mental_tidy %>%
  mutate(
    phq9_score = as.numeric(phq9_score),
    gad7_score = as.numeric(gad7_score)
  )

glimpse(mental_tidy)
```

```{r}
#let's inspect if there's anywhere in the dataset that numeric values are stored as text
mental_tidy %>%
  select(workhours, sleep, stressscale) %>%
  distinct() %>%
  head(10)
#We can see it's incosistently formatted (sometimes 'hours' sometimes 'h')


```

```{r}
mental_tidy <- mental_tidy %>%
  mutate(
    workhours = as.numeric(str_extract(workhours, "\\d+\\.?\\d*")),
    sleep = as.numeric(str_extract(sleep, "\\d+\\.?\\d*"))
  )

mental_tidy %>%
  select(workhours, sleep) %>%
  head(10)
```

```{r}
#Was getting all NA so ran everything again in one code chunk: 

mental_tidy <- mental_clean %>%
  pivot_longer(
    cols = jan:may,
    names_to = "month",
    values_to = "monthly_value"
  ) %>%
  mutate(across(where(is.character), ~ na_if(.x, "missing"))) %>%
  mutate(across(where(is.character), ~ na_if(.x, "-"))) %>%
  mutate(across(where(is.character), ~ na_if(.x, "n/a"))) %>%
  mutate(
    phq9_score = str_extract(screeningscores, "(?<=PHQ-9=)\\d+|(?<=PHQ9:)\\d+"),
    gad7_score = str_extract(screeningscores, "(?<=GAD-7=)\\d+|(?<=GAD7:)\\d+")
  ) %>%
  mutate(
    phq9_score = as.numeric(phq9_score),
    gad7_score = as.numeric(gad7_score)
  )

mental_tidy %>%
  filter(!is.na(workhours) | !is.na(sleep) | !is.na(stressscale)) %>%
  select(workhours, sleep, stressscale) %>%
  slice(1:10)
```

```{r}
mental_tidy <- mental_tidy %>%
  mutate(
    workhours = as.numeric(str_remove_all(workhours, "[^0-9.]")),
    sleep = as.numeric(str_remove_all(sleep, "[^0-9.]")),
    stressscale = as.numeric(stressscale)
  )

mental_tidy %>%
  filter(!is.na(workhours) | !is.na(sleep) | !is.na(stressscale)) %>%
  select(workhours, sleep, stressscale) %>%
  slice(1:10)
```

```{r}
#Final tidy dataset check

glimpse(mental_tidy)
```

At this point, the dataset has been transformed into a tidy structure. Monthly observations are stored in a single month column, screening scores have been separated into individual variables, and numeric values previously stored as text have been cleaned and converted for analysis.

Let's begin our analysis.

```{r}
#Average screening scores by age group

mental_summary <- mental_tidy %>%
  group_by(agegroup) %>%
  summarise(
    avg_phq9 = mean(phq9_score, na.rm = TRUE),
    avg_gad7 = mean(gad7_score, na.rm = TRUE)
  )

mental_summary
```

Let's make the above into a chart for visualizing.

```{r}
mental_summary_long <- mental_summary %>%
  pivot_longer(
    cols = c(avg_phq9, avg_gad7),
    names_to = "score_type",
    values_to = "average_score"
  )

ggplot(mental_summary_long, aes(x = agegroup, y = average_score, fill = score_type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average PHQ-9 and GAD-7 Scores by Age Group",
    x = "Age Group",
    y = "Average Score",
    fill = "Score Type"
  )
```

This summary compares average depression screening scores (PHQ-9) and anxiety screening scores (GAD-7) across age groups. Grouping by age category helps show whether mental health burden differs across demographic segments. It's clear that 18-25, 36-45 and 46-60 generally have the highest PHQ-9 scores while 36-45 and unknown age groups have the highest GAD-7 scores.
