---
title: "Week 3B Codebase"
author: "Theresa Benny"
format: html
editor: visual
---

## Approach

I'll be analyzing time-series data for at least two separate items and use **window functions** to compute:

-   **Year-to-Date (YTD) average**

-   **Six-day moving average**

I would focus not just on calculating averages, but on applying window functions correctly within grouped time-series data.

Dataset Selection

I will use a time-series dataset containing daily closing prices for at least two financial instruments (e.g., two stocks or cryptocurrencies) starting from January 1, 2022.

The dataset will include:

-   `date`

-   `item` (stock or cryptocurrency name/symbol)

-   `closing_price`

This structure allows grouping by item and ordering by date which are both required for window functions.

Data Preparation

Before applying window functions, I will:

1.  Ensure the `date` column is stored as a proper date type.

2.  Sort the dataset by `item` and `date`.

3.  Verify no duplicate dates per item.

4.  Handle missing dates or values if necessary.

The data will be grouped by `item` before calculating rolling or cumulative metrics.

Year-to-Date (YTD) Average

The YTD average represents the cumulative average price for each item from January 1 of the current year up to each specific date.

Using window logic:

-   Partition by `item`

-   Order by `date`

-   Use a cumulative window frame from the start of the year to the current row

Six-Day Moving Average The six-day moving average calculates the average closing price over the current day and the previous five days.

Using window logic:

-   Partition by `item`

-   Order by `date`

-   Use a rolling window frame of 6 rows (current row + 5 preceding rows)

For the first five days (where fewer than six observations exist), the result will either:

-   Be `NA`, or use partial windows (depending on implementation choice)

**Anticipated Challenges**

1\. Ordering and Grouping

Window functions depend heavily on correct ordering. Any sorting errors will produce incorrect rolling averages.

2\. Year Boundaries

If multiple years are included, YTD must reset each January 1. This requires partitioning by both item and year. 

##Codebase

```{r}
library(dplyr)
library(lubridate)
library(tibble)

#LLM generated the below dataset for Apple and Microsoft stocks
stock_data <- tibble(
  date = rep(seq.Date(as.Date("2024-01-02"), by = "day", length.out = 10), 2),
  stock = rep(c("AAPL", "MSFT"), each = 10),
  close = c(
    # Apple closing prices (realistic Jan 2024 levels)
    185, 187, 186, 188, 190, 191, 189, 192, 193, 195,
    # Microsoft closing prices
    370, 372, 371, 374, 376, 378, 377, 379, 381, 382
  )
)

stock_data

```
```{r}
#Now calculate YTD average using the cummean function in dplyr
stock_ytd <- stock_data %>%
  group_by(stock, year(date)) %>%
  arrange(date) %>%
  mutate(
    ytd_avg = cummean(close)
  )

stock_ytd

```

```{r}
#this library has the rollmean function to calculating the rolling-average
library(zoo)

stock_ma <- stock_ytd %>%
  group_by(stock) %>%
  arrange(date) %>%
  mutate(
    ma_6day = rollmean(close, k = 6, fill = NA, align = "right")
  )

stock_ma

```
```{r}
stock_ma %>%
  filter(stock == "AAPL") %>%
  select(date, close, ytd_avg, ma_6day)

library(gt)

stock_ma %>%
  select(stock, date, close, ytd_avg, ma_6day) %>%
  gt(groupname_col = "stock") %>%
  tab_header(
    title = "Daily Closing Prices with YTD and 6-Day Moving Average by Stock"
  ) %>%
  cols_label(
    stock   = "Stock",
    date    = "Date",
    close   = "Closing Price",
    ytd_avg = "YTD Average",
    ma_6day = "6-Day Moving Avg"
  ) %>%
  fmt_number(
    columns = c(close, ytd_avg, ma_6day),
    decimals = 2
  )

```

